
---
output: html_document
---
Red Wine Quality EDA by Adrien Viani
========================================================
#Citation Request
>  This dataset is public available for research. The details are described in [Cortez et al., 2009]. 
  Please include this citation if you plan to use this database:  
  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.   
  Modeling wine preferences by data mining from physicochemical properties.  
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.  
  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016  
                [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf  
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib  
                


```{r echo=FALSE, message=FALSE, warning=FALSE}
# sets the global heights and widths for plots
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

# load packages for analysis
library(ggplot2)
library(gridExtra)
library(GGally)
library(grid)
library(scales)
library(memisc)
library(plotly)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Graph_Function}

# defining a function to graph a box plot and histogram / bar chart side by side
# function takes in a variable and string name, and bin length
graph_fun <- function(dat.var, dat.var.name = '', binwidth = NULL) {
  # creates a histogram with default bar size, blue bars and black outlines
  p1 <- ggplot(aes(x = dat.var), data = dat) +
  geom_histogram(binwidth = binwidth, fill = 'blue', colour = 'black') +
  xlab(dat.var.name)
  
  # Creates a boxplot with points overlayed with jitter and alpha, to better
  # display density 
  p2 <- ggplot(aes(x="", y =dat.var), data = dat) +
    geom_boxplot(outlier.shape = NA)+
    geom_jitter(alpha = 1/6, shape=16, position=position_jitter(0.1), color = 'darkblue')+
        ylab(dat.var.name)+
    xlab("")
  coord_flip()

# arranges the plots in a 1x2 grid  
grid.arrange(p1,p2, ncol=2)

}
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
#set the working directory, read in the data
setwd('C:/Users/avian/Desktop/Red Wine EDA Project')
dat <- read.csv("wineQualityReds.csv", header = TRUE)
```

#Description of Paper and Data
> This EDA paper focuses on the effect of various red wine characteristics on 
perceived quality. The dataset and information related to it is available in the
citation above. The 12 variables explored in this paper are listed below with
a brief description.
<br/><br/>
Description of attributes:    
1 - fixed acidity: most acids involved with wine or fixed or nonvolatile (do not
evaporate readily)  
2 - volatile acidity: the amount of acetic acid in wine, which at too high of le
vels can lead to an unpleasant, vinegar taste  
3 - citric acid: found in small quantities, citric acid can add 'freshness' and 
flavor to wines  
4 - residual sugar: the amount of sugar remaining after fermentation stops, it's
rare to find wines with less than 1 gram/liter and wines with greater than 45 gr
ams/liter are considered sweet  
5 - chlorides: the amount of salt in the wine  
6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between mole
cular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth a
nd the oxidation of wine  
7 - total sulfur dioxide: amount of free and bound forms of S02; in low concentr
ations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 
50 ppm, SO2 becomes evident in the nose and taste of wine  
8 - density: the density of water is close to that of water depending on the per
cent alcohol and sugar content  
9 - pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
to 14 (very basic); most wines are between 3-4 on the pH scale  
10 - sulphates: a wine additive which can contribute to sulfur dioxide gas (S02)
levels, wich acts as an antimicrobial and antioxidant  
11 - alcohol: the percent alcohol content of the wine  
Output variable (based on sensory data): 
12 - quality (score between 0 and 10)
<br/><br/>
Units of Attributes:  
1 - fixed acidity (tartaric acid - g / dm^3)  
2 - volatile acidity (acetic acid - g / dm^3)  
3 - citric acid (g / dm^3)  
4 - residual sugar (g / dm^3)  
5 - chlorides (sodium chloride - g / dm^3)  
6 - free sulfur dioxide (mg / dm^3)  
7 - total sulfur dioxide (mg / dm^3)  
8 - density (g / cm^3)  
9 - pH  
10 - sulphates (potassium sulphate - g / dm3)  
11 - alcohol (% by volume)  
Output variable (based on sensory data):   
12 - quality (score between 0 and 10)  


# Univariate Plots Section
> In this section, I will plot a boxplot and histogram for each variable laid out
above, discuss the qualities of the distribution, and possibly log transform the
variables to better approximate a normal distribution. 

```{r echo=FALSE, message=FALSE, warning=FALSE }
# print 5 num summ., use graphing function for box plot and histogram of quality
summary(dat$quality)
table(dat$quality)
graph_fun(dat$quality, "quality", 1)
```

> Quality appears roughly normally distributed. Of note is that in 
our dataset it appears that the bulk of the results fall within the range 5-6. 
The relative sparseness of low quality and high quality wines in the sample
could present a balancing issue in any predictive analysis pursued down the line.


```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$fixed.acidity)
graph_fun(dat$fixed.acidity, "Fixed Acidity")
```

> Fixed acidity is approximately normal with a slight right
skew and potential outliers at the end of the tail. The bulk of the falls betwee
n 7.1-9.2 (tartaric acid - g / dm^3)


```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$volatile.acidity)
graph_fun(dat$volatile.acidity, "Volatile Acidity")
```

> Volatile acidity also appears approximately normal with a slight right skew an
d potential outliers at the end of the tail. The bulk of the results fall betwee
n .39 and .64   (acetic acid - g / dm^3)


```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$citric.acid)
graph_fun(dat$citric.acid, "Citric Acid")
```

> Citric acid does not appear to be normally distributed. There are peaks in frequency at 0, and 0.5, with other less stark peaks at approximately .14 and .2 (g / dm^3)


```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$residual.sugar)
graph_fun(dat$residual.sugar, "Residual Sugar")
```

> Residual sugar is right skewed with a very long tail. The bulk of residual 
sugar data lies between .27 and .36 (g / dm^3). After a log transform,shown 
below, it appears to be closer to normally distributed albeit with some right 
skew remaining.



```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(log10(dat$residual.sugar))
graph_fun(log10(dat$residual.sugar), "Log10(Residual Sugar)")
```



```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$chlorides)
graph_fun(dat$chlorides, "Chlorides")
```

> Chlorides is right skewed with a very long right tail. The bulk of chlorides d
ata lies between .07 and .09 (g / dm^3). After a log transform,shown below, it 
appears to be closer to normally distributed albeit long, skinny tails


```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(log10(dat$chlorides))
graph_fun(log10(dat$chlorides), "Log10(Chlorides)")
```



```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$free.sulfur.dioxide)
graph_fun(dat$free.sulfur.dioxide, "Free Sulfur Dioxide")
```

> Free sulfur dioxide is right skewed. The bulk of the data lies between 
7 and 21 (g / dm^3). After a log transform it does not appear normally 
distributed, and actually looks bimodal.

```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(log10(dat$free.sulfur.dioxide))
graph_fun(log10(dat$free.sulfur.dioxide), "Log10(Free Sulfur Dioxide)")
```

```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$total.sulfur.dioxide)
graph_fun(dat$total.sulfur.dioxide, "Total Sulfur Dioxide")
```

> Total sulfur dioxide is right skewed with a very long right tail. The bulk of 
total sulfur dioxide data lies between .07 and .09 (mg / dm^3). After a log 
transform,shown below, it appears to be closer to normally distributed.


```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(log10(dat$total.sulfur.dioxide))
graph_fun(log10(dat$total.sulfur.dioxide), "Log10(Total Sulfur Dioxide)")
```


```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$density)
graph_fun(dat$density, "Density")
```

> Density appears normally distributed, with a mean of .9967 and the bulk of 
results falling between .9956 and .9968 (g / cm^3)


```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$pH)
graph_fun(dat$pH, "pH")
```

> pH appears normally distributed, with a mean of 3.311 and the bulk of 
results falling between 3.21 and 3.4. 

```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$sulphates)
graph_fun(dat$sulphates, "Sulphates")
```

 > Sulphates is right skewed. The bulk of sulphate data lies between .55 and 
.73 (potassium sulphate - g / dm3). After a log transform,shown below, 
it appears to be closer to normally distributed.


```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(log10(dat$sulphates))
graph_fun(log10(dat$sulphates), "Log10(Sulphates)")
```



```{r echo=FALSE, message=FALSE, warning=FALSE, }
summary(dat$alcohol)
graph_fun(dat$alcohol, "Alcohol")
```

> Finally alcohol does not appear to be normally distributed. Alcohol is slightly
right skewed, but does not look like it would benefit from a log transform. 
The bulk of the results fall between 9.5 and 11.10 (% by volume)

# Univariate Analysis

### What is the structure of your dataset?

>The dataset covers 1,599 observations across 12 variables. All variables are 
real, continous numbers except for quality which is factor (integer) levels 0 (low) 
through 10 (high).
<br\>
Of note is the fact that multiple values of quality are not present in the 
dataset (0, 1, 2, 9, and 10). Other variables present are related by definition 
(see above units) and will likely have high covariance with each other.

### What is/are the main feature(s) of interest in your dataset?

> The primary feature for my analysis will be quality, but I will also explore 
some of the covariance of other properties related by unit definition and as 
indicated by a correlation plot

### What other features in the dataset do you think will help support your analysis?
> I believe alcohol, citric acid, sulphates, and sugar will support my analysis 
as these are the qualities I immediately recognize from my knowledge of wine.
I am sure thought that in exploring the data I may discover new connections I 
wasn't aware of before.

### Did you create any new variables from existing variables in the dataset?

>I did not create any new (composite) variables from the dataset, but did log
transform some of the data in my initial exploration.

###Of the features you investigated, were there any unusual distributions?
###Did you perform any operations on the data to tidy, adjust, or change the form of the data? \ 
###If so, why did you do this?

> When exploring the data I log transformed a couple of skewed distributions
which included residual sugar, chlorides, free sulfur dioxide, and total sulfur
dioxide. The log transform ideally reduces tail size / impact which can affect 
regression analysis in the future.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
# correlation plot, used to pick which variables have strongest correlations
# to explore further down
ggcorr(data = dat)
```

> We can see from this correlation plot that there are many strong correlations
that can be explored. We will graph and discuss some of these pairs as listed 
below.
<br\>
Density: Residual Sugar, Fixed Acidity, Citric Acid, Alcohol<br\>
Fixed Acidity: Citric Acid, Volatile Acidity<br\>
Free Sulfur Dioxide, Total Sulfur Dioxide<br\>
Quality: Alcohol, Sulphates, Volatile Acidity, Citric Acid, Residual Sugar


```{r echo=FALSE, message=FALSE, warning=FALSE,}
# the sections below create two way graphs of the above listed variable pairs
# each graph displays a linear smoother to give an idea of "trend-line" 
ggplot(aes(x = density, y = residual.sugar), data = dat) +
  geom_point(alpha = 1/2, position = position_jitter()) +
  stat_smooth(method = 'lm')
```

> There is a mild positive correlation between density and residual sugar

```{r echo=FALSE, message=FALSE, warning=FALSE, }
ggplot(aes(x = density, y = fixed.acidity), data = dat) +
  geom_point(alpha = 1/2, position = position_jitter()) +
  stat_smooth(method = 'lm')

```

> There is a moderate positive correlation between desnity and fixed acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = density, y = citric.acid), data = dat) +
  geom_point(alpha = 1/2, position = position_jitter()) +
  stat_smooth(method = 'lm')

```

> Based on the correlation chart above, there is a mild positive correlation
between citric acid and density. From the chart, we can see thought that this 
relationship is quite noisy.

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = density, y = alcohol), data = dat) +
  geom_point(alpha = 1/2, position = position_jitter()) +
  stat_smooth(method = 'lm')
```

> Finaly there is a moderate negative correlation between density and alcohol.
This makes sense from a chemical perspective as alcohol is less dense than water.


```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = fixed.acidity, y = citric.acid), data = dat) +
  geom_point(alpha = 1/2, position = position_jitter()) +
  stat_smooth(method = 'lm')
```

> There is a moderate-strong correlation beetween fixed acidity and volatile 
acidity although the data is quite noisy.

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = fixed.acidity, y = volatile.acidity), data = dat) +
  geom_point(alpha = 1/2, position = position_jitter()) +
  stat_smooth(method = 'lm')
```


```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = volatile.acidity, y = citric.acid), data = dat) +
  geom_point(alpha = 1/2, position = position_jitter()) +
  stat_smooth(method = 'lm')
```

> There is a slight negative correlation between volatile acidity and citric aci
d but the data is quite noisy.

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide), data = dat) +
  geom_point(alpha = 1/2, position = position_jitter()) +
  stat_smooth(method = 'lm')
```

> There is a moderate positive correlation between free and total sulfur dioxide
, which makes some degree of intuitive sense given that the measures are 
related

```{r echo=FALSE, message=FALSE, warning=FALSE,}
# each of the box plots below explores a variable at differing quality levels
# this provides an alternative visual means of identifying relevant trends
ggplot(aes(x = quality, y =alcohol, group = quality), data = dat) +
  geom_boxplot()
```

> Quality is clearly increasing with alcohol %.

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = quality, y =sulphates, group = quality), data = dat) +
  geom_boxplot() 
```

> Quality is mildly increasing with sulphates, although it is clear that there 
are outliers in sulphates at each quality level, particularly quality 5-6. This
could be indicative that extremely high values of sulphates (above a particular
threshold) could potentially be a negative influence on wine quality.

```{r echo=FALSE,message=FALSE, warning=FALSE,}
ggplot(aes(x = quality, y = volatile.acidity, group = quality), data = dat) +
  geom_boxplot()
```

> Quality increases as volatile acidity decreases. This interaction makes sense 
as volatile acidity is what imparts a vinegar taste to wine.

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = quality, y = citric.acid, group = quality), data = dat) +
  geom_boxplot()
```

> Quality is increasing with citric acid. This interaction makes sense as citric
acid is perceived to add "freshness" and a pleasant taste to wines.

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = quality, y = residual.sugar, group = quality), data = dat) +
  geom_boxplot()
```

> Quality is flat with residual sugar, although it is clear that there are 
outliers at each qualty level. It appears that the highest (and generally 
densest) sugar outliers are in the middle of the quality range, which may 
represent an insight  into flavor preferences for high quality red wines 
(ie. a preference for "dry"wines)

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = quality, y = free.sulfur.dioxide, group = quality), data = dat) +
  geom_boxplot()
```

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(aes(x = quality, y = total.sulfur.dioxide, group = quality), data = dat) +
  geom_boxplot()
```

> There appears to be a mildly parabolic relationship between quality and both 
free and total sulfur dioxide, with larger tails for each of the sulfur measures
in the middle quality range (5-7).

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

> Quality is positively correlated with alcohol, citric acid, and sulphates 
while it is negatively correlated with volatile acidity. Interestingly enough, 
residual sugar is not strongly correlated with quality, but it appears that the
middle of the quality range has the largest high sugar outliers (ie. larger 
tails). This could indicate that overly sweet wine may not rate as well as 
similar wines that have less sugar. Quality is mildly parabolic with both sulfur 
measures, with similar tail behavior as sugar in the 5-7 range.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

> There was nothing terribly surprising in the elements that I investigated, but
it was reassuring to confirm certain relationships that made chemical sense
based on variable descriptions ie. density vs alcohol and total vs free sulfur 
dioxide. It might be interesting to investigate whether more sugar (holding 
other positive traits equal like alcohol, citric acid, and possibly sulphates) 
has a potentially negative relationship with quality.

### What was the strongest relationship you found?

> Alcohol and density, is a strong relationship primarily due to the chemical 
properties of alcohol. As for quality, citric acid and alcohol look strongly 
related to quality.


# Multivariate Plots Section



```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
# plot variables against each other in a scatterplot
# this time though, quality is used to create a color gradient in the points
# this can help identify trends visually
ggplot(dat, aes(x=alcohol, y = citric.acid, color = as.factor(quality)))+
  geom_point()+
  scale_color_brewer(type = "div", palette = "RdYlBu", name="Quality", direction=-1)

```

> In plotting alcohol percentage and citric acid together with quality as a 
color gradient, it is clear that more of the high quality wines fall in high 
alcohol and high citric acid regions.

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(dat, aes(x=alcohol, y = sulphates, color = as.factor(quality)))+
  geom_point()+
  scale_color_brewer(type = "div", palette = "RdYlBu", name="Quality", direction=-1)
```

> In plotting alcohol percentage and sulphates together with quality as a 
color gradient, it is clear that more of the high quality wines fall in high 
alcohol regions, while sulphates appears to have less of a linear relation with 
quality and more of a parabolic one. (Quality seems to drop off, outside of a 
specific band of sulphate quantity)

```{r echo=FALSE, message=FALSE, warning=FALSE,}
ggplot(dat, aes(x=volatile.acidity, y = free.sulfur.dioxide, color = as.factor(quality)))+
  geom_point()+
  scale_color_brewer(type = "div", palette = "RdYlBu", name="Quality", direction=-1)
```

> Plotting volatile acidity with free sulfur dioxide with quality as a color 
gradient shows declining quality with increases of volatile acidity, less so 
free sulfur dioxide. It does appear though that wines can still rank 7+ even with 
high(er) free sulfur dioxide. From the bivariate plot above, we do see that free
and total sulfur dioxide are positively related, it is likely that better wines 
with high free sulfur dioxide fall on the lower range of total sulfur in the 
sample for those particular free sulfur levels.



```{r echo=FALSE, message=FALSE, warning=FALSE,}

dat$quality.fac <- as.factor(dat$quality)

# plotly 3d plot to explore residual sugar and quality interaction across wines
# of similar alcohol and citric acid characteristics
p <- plot_ly(dat, x= ~ alcohol, y = ~citric.acid, z = ~residual.sugar, 
             opacity = 0.5, color = ~quality.fac, colors = 'RdYlBu',
             text= ~paste('Quality: ', quality), hoverinfo = "x+y+z+text") %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'X: % Alcohol'),
                      yaxis = list(title = 'Y: Citric Acid'),
                      zaxis = list(title = 'Z: Residual Sugar')))

p
```
> Here we have a 3d graph that examines the impact sugar has on quality
for different levels of citric acid and alcohol content. It is clear that for 
similar levels of citric acid and alcohol content, increasing sugar content 
appears correlated with lower wine quality. Because the bulk of the sample falls
in the "middle" range of quality 5-7, it would be good to have a broader sample
to see if this holds across "slices" of citric acid and alcohol.


```{r echo=FALSE, message=FALSE, warning=FALSE,}

# plotly 3d plot to explore sulphates and quality interaction across wines
# of similar alcohol and citric acid characteristics
g <- plot_ly(dat, x= ~ alcohol, y = ~citric.acid, z = ~sulphates, 
             opacity = 0.5, color = ~quality.fac, colors = 'RdYlBu', 
             text= ~paste('Quality: ', quality), hoverinfo = "x+y+z+text") %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'X: % Alcohol'),
                      yaxis = list(title = 'Y: Citric Acid'),
                      zaxis = list(title = 'Z: Sulphates')))

g
```

> In this graph, using turntable rotation to isolate "faces" of the graph,
we can compare sulphate's interactions with alcohol, citric acid, and quality. 
It looks like there is a sulphates band, within which quality ratings are 
moderately higher. When the sulphates are higher or lower than that band, 
quality tends to decline all else equal.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

>It is clear that the variables that correlate positively with quality (citric 
acid, alohol) strengthened each other from the first plot in the multivarite 
section. The negative correlation variables examined did not (free sulfur 
dioxide and volatile acidity.)

> It also appears that higher residual sugar may contribute to declining quality
compared to wines with similar citric acid and alcohol levels but lower sugars.
Additionally, while sulphates appeared slightly positively correlated in the 2d
boxplot of sulphates vs quality, the 3d plot appears to indicate that it is more
likely that there is an "optimal" range of sulphates in wine. Above or below 
this range corresponds with a drop off in quality.

### Were there any interesting or surprising interactions between features?

> I think the residual sugar and sulphate properties discussed above are 
interesting. It points to the necessity of careful interpretation and 
exploration before drawing conclusions to best understand covariates. This is 
especially important before performing any modeling as it could lead to 
incorrect or incomplete understanding.

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE,}
# this code arranges what I perceived to be the most meaningful boxplots of
# variables against quality in a 1x4 grid
p1 <- ggplot(aes(x = quality, y = alcohol, group = quality), data = dat) +
  geom_boxplot(outlier.shape =NA)+
  geom_jitter(alpha = 1/8, shape=16, position=position_jitter(0.1), color = 'blue')+
  xlab('Quality')+
  ylab('Alcohol (% of volume)')

p2 <- ggplot(aes(x = quality, y = sulphates, group = quality), data = dat) +
  geom_boxplot(outlier.shape =NA) +
  geom_jitter(alpha = 1/8, shape=16, position=position_jitter(0.1), color = 'blue')+
  xlab('Quality')+
  ylab('Sulphates (potassium sulphate - g/dm³)')

p3 <- ggplot(aes(x = quality, y = citric.acid, group = quality), data = dat) +
  geom_boxplot(outlier.shape =NA)+
  geom_jitter(alpha = 1/8, shape=16, position=position_jitter(0.1), color = 'blue')+
  xlab('Quality')+
  ylab('Citric Acid (g/dm³)')

p4 <- ggplot(aes(x = quality, y = volatile.acidity, group = quality), data = dat) +
  geom_boxplot(outlier.shape =NA)+
  geom_jitter(alpha = 1/8, shape=16, position=position_jitter(0.1), color = 'blue')+
  xlab('Quality')+
  ylab('Volatile Acidity (acetic acid - g/dm³)')


grid.arrange(p1,p3,p2,p4, ncol=4, top = paste('Boxplots of Alcohol, Sulphates,',
'Citric Acid, and Volatile Acidity By Quality', sep = ' '))
```

### Description One

> This set of four plots shows some of the relationships between quality and 
variables in the data set. We note positive correlations with alcohol, citric 
acid and sulphates; we also note a negative correlation with volatile acidity. 
These plots are interesting as they are clearly visible interactions amongst
some of the 11 variables vs quality.


### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
# a cleaned up version of the plotly 3d scatter above
g <- plot_ly(dat, x= ~ alcohol, y = ~citric.acid, z = ~residual.sugar, 
             opacity = 0.5, color = ~quality.fac, colors = 'RdYlBu',
             text= ~paste('Quality: ', quality), hoverinfo = "x+y+z+text") %>%
  add_markers() %>%
  layout(title = paste('Plot of Alcohol, Citric Acid, Residual Sugar',
                       'and Quality (using color)',sep=' '),
         scene = list(xaxis = list(title = 'X: Alcohol (% By Volume)'),
                      yaxis = list(title = 'Y: Citric Acid (g/dm³)'),
                      zaxis = list(title = 'Z: Residual Sugar (g/dm³)')))

g
```

### Description Two

> This graph examines the interaction of sugar, alcohol and citric acid with 
quality. In examining the box plots of sugar vs quality, I noticed that high
sugar outliers were more common in the 5-7 quality range, and could affect 
quality negatively. To explore this, I needed a plot where I could compare and
contrast wines with similar citric acid and alcohol profiles with differing
varying levels of sugar. While the data is very clustered, and sampling at both 
the top and bottom of the quality range is low, it does appear that higher sugar
correlates with lower ratings within similar citric acid and alcohol profiles.
This is interesting because it indicates it might be worth exploring various 
thresholds to define new factor variables where sugar is high, to improve and
explore predictive modeling down the road.




### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
# a cleaned up version of the plotly 3d scatter above
g <- plot_ly(dat, x= ~ alcohol, y = ~citric.acid, z = ~sulphates, 
             opacity = 0.5, text= ~paste('Quality: ', quality), hoverinfo = "x+y+z+text") %>%
  add_markers(color = ~quality.fac, colors = 'RdYlBu')%>%
  layout(title = paste('Plot of Alcohol, Citric Acid, Sulphates',
                       'and Quality (using color)',sep=' '),
         scene = list(xaxis = list(title = 'X: Alcohol (% By Volume'),
                      yaxis = list(title = 'Y: Citric Acid (g/dm³)'),
                      zaxis = list(title = 'Z: Sulphates (potassium sulphate - g / dm³)')))

g
```

### Description Three

> This graph examines sulphates, citric acid, and alcohol and their combined 
interactions on quality. By turning this graph using the turntable function in 
the interactive menu, one can look at different angles and highlight various 
points to better understand the interaction of sulphates with quality and the 
other variables present. By moving the graph into a "2d" visual, one can see 
that there appears to an optimal range for sulphates to achieve good quality. 
Outside of this band, wine quality appears to drop off across all alcohol and 
citric acid ranges. This shows that while the initial boxplot may have indicated
a positive correlation between sulphates and quality, that the actual 
relationship may be slightly more complex.


# Reflection

> I enjoyed exploring the red wine dataset, and found learning R to be relatively
straightforward. I found the data mostly easy to work with, and was able to 
learn about functions in R to expedite some of my work. I also learned more 
about the plotly package, and was able to exploit the 3D scatterplot in the
package to better explore how sulphates and residual sugar affect quality in the
sample. Based on feedback in my project review, I was able to improve the 
graphing of factor levels in quality, and greatly improved the presentation of 
my multivariate graphs by setting discrete color levels with sensible color 
palettes.
<br\>

> One of the challenges of the dataset was interpreting data that didn't have 
clear positive and negative correlations with quality, like sugar and sulphates.
The experience emphasized the importance of looking beyond common basic measures
like quartiles and means. I could see 3d graphing and examining tail behavior as
an important part of my future EDA and modeling efforts. I also found this 
portion of the analysis challenging, as there is limited data at the high and 
low end of the quality ranges, making it hard to visualize certain interactions.
<br\>

### Idea for Future Work
> In the future, I would like to explore predicting wine quality using basic 
classification models. I could split quality levels into various buckets (ie.
good, average, bad) to solve the balancing issues. From there I could balance 
the data by using standard sampling methods. Additionally, I could transform
some of the variables to better quantify identified behavior. For example, 
creating a centered and squared version of sulphates could better identify the
"optimal quality range" behavior I identified. Similarly a transformation that
seperates sugar outliers at similar levels of citric acid and alcohol may be 
useful. After that work, I could apply a random forest based method to determine
the importance of the variables in the dataset and further my analysis.
